### 2021-12-22 Manuscript.Rmd
FZ, SS, CWD comments

---
bibliography: references.bib
csl: council-of-science-editors-author-date.csl
site: "bookdown::bookdown_site"
header-includes:
    - \usepackage{setspace}\doublespacing

output:
  bookdown::pdf_document2:
    toc: FALSE
    fig_caption: yes
---

@@

```{r zooids-definition, echo=FALSE, message=F, warning=FALSE, verbose= FALSE, out.height='0.5\\textheight', out.width='0.5\\textheight', fig.cap='(ref:zooids-definition)', fig.align = 'center'}

@@

Using STF and SBF, combined with classical differential gene expression (DGE) analysis, we conducted DGE analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. We also examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Second, we compared gene expression between species, using STF followed by linear models to identify zooid- and species-variable genes. And third, we used phylogenetic methods and SBF to identify putative clade/lineage or zooid specific expression patterns.

@@

```{r methods-comp, echo=FALSE, message=F, warning=FALSE, verbose= FALSE, out.height='0.7\\textheight', out.width='0.7\\textheight', fig.cap='(ref:methods-comp)', fig.align = 'center'}

@@

The third component of variation we considered was among zooids/specialized tissues within species. This was the greatest component of variation (Fig. \@ref(fig:pca-diphyes)--\@ref(fig:pca-physalia)). We identified genes that significantly differentially expressed in particular zooids, and in the pneumatophore, for each of the sampled species, based on pairwise comparisons (supplementary file 1). In each case, significally differentially expressed refers to genes with higher transcript abundance relative to the other zooid. As it is possible for the same gene to be significantly differentially expressed in pairwise comparisons of several zooids , we also identified a subset of genes that are significantly differentially expressed in only one zooid but not in any other zooid within a given species (Fig. \@ref(fig:unique-zooids), supplementary file 2).

@@

In siphonophores, there are several instances of lineage-specific zooid diversification events. We investigated gene expression patterns between the novel zooid type and the hypothesized most closely related zooid type in three species. In *Bargmannia elongata* there are two morphologically distinct gastrozooids, that we termed "white" and "yellow" gastrozooids (Fig. \@ref(fig:novel-zooids)A and \@ref(fig:novel-zooids)B). The "yellow" gastrozooid is larger and darker and occurs as the 7th-10th gastrozooid on the stem [@dunn2005complex].  In the Portuguese man of war, *Physalia physalis*, the gastrozooid is unique compared to other gastrozooids in other species - it has a mouth, but no tentacle, and the basigaster region is greatly reduced [@Totton1960; @Mackie1960]. Meanwhile the tentacle is associated with another zooid, the tentacular palpon (Fig. \@ref(fig:novel-zooids)C) [@BARDI2007;@munro2019morphology; @Totton1960; @haeckel1888]. In *P. physalis*, both the gastrozooid and the tentacular palpon are considered to be subfunctionalized from an ancestral gastrozooid type [@munro2019morphology]. Finally, in *Agalma elegans*, there are thought to be at least two different palpon types: gastric palpons that arise at the base of the peduncle of the gastrozooid, and a palpon called the B-palpon (Fig. \@ref(fig:novel-zooids)D) [@dunn2006evolution]. The distinction between these two types of palpon is based on the location of these zooids - the gastrozooid is typically the last element of each cormidium, but based on the budding sequence, Dunn *et al* propose that the enlarged B-palpon is the last element in *A. elegans* [@dunn2006evolution]. Each of these cases represents a different type of novelty: in *Bargmannia elongata* the distinction between zooids was made based on size and color but not on obvious differences in function, in *P. physalis* the gastrozooids and tentacular palpons differ structurally and functionally, and finally in *A. elegans*, gastric palpons and B palpons differ only in colony location, development, and possibly size.

@@

pca_data2<-prcomp(t(pcatest2))
pca_data_perc2<-round(100*pca_data2$sdev^2/sum(pca_data2$sdev^2),1)
test_names2<-colnames(pcatest2) %>% str_split("_")
species2<-lapply(test_names2, `[[`, 1) %>% unlist() %>% str_replace("\\."," ")
treatment2<-lapply(test_names2, `[[`, 2) %>% unlist() 
#replicate<-lapply(test_names, `[[`, 3) %>% unlist()
df_pca_data2 <- data.frame(PC1 = pca_data2$x[,1], PC2 = pca_data2$x[,2], species=species2, treatment=treatment2)
plot_pca_ortholog2<-df_pca_data2 %>% ggplot()+geom_point(aes(PC1,PC2, col=treatment, shape=species), size = 3) + 
  scale_shape_manual(values=c("Agalma elegans"=15,"Nanomia bijuga"=0,"Frillagalma vityazi"=16,"Bargmannia elongata"=9)) +
  xlab(paste0("PC1: ",pca_data_perc[1],"% variance")) +  ylab(paste0("PC2:",pca_data_perc[2],"% variance")) +
  scale_color_manual(values=c("GasdevGasmat"="#F78508","NecdevGasmat"="#B9C977","PalmatGasmat"="#F4DF1C","PneGasmat"="#8DCCE2"))

@@

(ref:SOFT-plot) Results of classical gene expression analyses across species, using Speciation Tree Filtering followed by linear models. Top row (A & B): results using TPM10K expression values; Bottom row (C & D): results using TPM10K expression ratios. For both analyses of expression variance explained by species and tissue (right-hand column), we identified highly species-variable genes (SVG, red) and highly tissue-variable genes (TVG, blue). In SVGs, the proportion of variance explained by species is two times greater than that explained by tissues; while in TVGs the proportion of variance explained by tissues is two times greater than that explained by species. All SVGs and TVGs are genes for which both species and tissue explain greater than 75% of variance. A. PCA plot of expression values (TPM10K) of `r res %>% nrow()` strict ortholog genes, showing a species dominated clustering pattern. B.  The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using TPM10K expression values. C. PCA plot of expression ratios of `r res2 %>% nrow()` strict ortholog genes, showing a tissue dominated clustering pattern.  D. The proportion of expression variance explained by species (x-axis) versus the proportion of expression variance explained by tissues (y-axis), using expression ratios. Abbreviations: Gasdev = developing gastrozooid, Necdev = developing nectophore, Palmat = mature palpon, Pne = pneumatophore, Gasmat = mature gastrozooid.

```{r SOFT-plot, results='asis', echo=FALSE, message=FALSE, warning=FALSE, verbose= FALSE, fig.width=7.5, fig.height=7, fig.cap='(ref:SOFT-plot)', fig.align='center'}

@@

Most comparative studies of gene expression focus exclusively on strict 1:1 orthologs, requiring an assessment of orthology across all species. We call this type of analysis Speciation Tree Filtering (STF), as it limits analyses to a subset of genes with very specific evolutionary histories. Following this approach, we used Orthofinder 2 (v2.4.0) [@emms2019orthofinder] to identify strict orthologs. For all seven species, we identified `r Orthologs1$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r test3 %>% nrow()` orthologs had expression data across all 7 species. In order to increase the number of genes for analysis, we focused on the four best sampled species (*Agalma elegans*, *Bargmannia elongata*, *Frillagalma vityazi*, and *Nanomia bijuga*). Using data from 4 taxa, we identified `r Orthologs2$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r stvariance %>% nrow()` orthologs had expression values in four zooids/tissues: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore.  Using TPM10K alone, we found that orthologs clustered largely by species rather than zooid/ tissue (Fig. \@ref(fig:SOFT-plot)A). Following @breschi2016gene (see methods), we used linear models to identify the proportion of variance that could be explained by species or zooid/tissue (Fig. \@ref(fig:SOFT-plot)B), and found that of `r stvariance  %>% nrow()` total orthologs, `r stvariance %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance %>% filter(class=="TVG") %>% nrow()` zooid/tissue-variable genes (TVG).

@@

For our Speciation Branch Filtering (SBF) analyses (see Fig. \@ref(fig:methods-comp) and methods), we used transcriptome and genome data from `r species_tree$tip.label %>% length()` cnidarian species to generate a total of `r gene_trees %>% length()` gene trees, of which `r gene_trees_ace_calibrated[!is.na(gene_trees_ace_calibrated)] %>% length()` gene trees passed filtering criteria (branch length threshold <2, less than 0.25 branches/tree with default branch lengths, one or more speciation events present in the tree, fewer than 0.3 nodes that have descendant nodes with the same species node ID, that we able to be time calibrated and finally that have a root depth <5). We used Orthofinder 2 [@emms2019orthofinder] to reconcile the species tree and the gene trees and annotate each gene tree node as either a speciation or duplication event. The number of genes represented in these gene trees is shown in Table \@ref(tab:summarize-libraries). The internal nodes on these gene trees represent `r nodes_ace %>% filter(is.na(species) & Ev=="S") %>% nrow()` speciation events and `r nodes_ace %>% filter(is.na(species) & Ev=="D") %>% nrow()` duplication events. Expression values (TPM10K) were mapped to the tips of the gene trees for each zooid/tissue separately, and ancestral values were inferred at internal nodes with maximum likelihood. We focused on expression in a subset of zooids and tissues that are common across the sampled species: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore. The distribution of expression changes along species-equivalent branches are shown in figure \@ref(fig:changes-branches).

As with the STF and linear model analyses, we investigated the impact of counting efficiency on our comparative analyses. We used the TPM10K values at the tips and nodes to calculate expression ratios and calculated changes in expression ratios across a branch for pairs of tissues, using the mature gastrozooid values as the denominator. The distribution of expression ratio changes along species-equivalent branches (branches in the gene tree that correspond to equivalent branches in the species tree, and are descended from speciation events) are shown in figure \@ref(fig:changes-branches2). We found that the variance of change across a given species-equivalent branch is considerably higher for raw TPM10K values as compared with ratios of expression (Figs. \@ref(fig:variance-structure) & \@ref(fig:variance-structure2)). We also identified the number of gene tree branches that had a negative (change <= -2), positive (change >= 2) or neutral (change > -2 and < 2) change (Fig.\@ref(fig:neg-pos)). The mean number of species-equivalent branches considered in these analyses are shown in Table \@ref(tab:summarize-branches) -- this value reflects the mean number of branches across each zooid or zooid ratio (due to incomplete sampling, some zooids may have fewer or more branches; as seen in Fig.\@ref(fig:neg-pos)). For each species-equivalent branch, we found more negative and positive branches than neutral branches when we used TPM10K values, while for expression ratios, we found that the majority of branches show no change across the branch. This suggests that as for the STF analyses, counting efficiency has a large impact on expression change and in turn leads to large gene/branch specific signal. For expression ratios, these results indicate that the vast majority of branches show close to 0 (neutral) change across the branch, suggesting that for closely related genes, expression ratios tend not to differ.

(ref:neg-pos) Results of phylogenetic analysis of gene expression analyses, using Speciation Branch Filtering. Number of branches with negative, positive or neutral changes across species-equivalent branches. Top panel: species phylogeny with branch IDs given as letters. Bottom left panel: changes of TPM10K expression across each of the species-equivalent branches (IDs correspond to branch IDs shown in the species tree); bottom right panel: changes of expression ratios across species equivalent branches. Neutral change (green) is defined as <= 2 and >= -2, negative change (red) is defined as <-2, and positive change (blue) is defined as > 2. These units correspond to TPM10K ratios/substitutions per site. GasdevGasmat = developing gastrozooid / mature gastrozooid, NecdevGasmat = developing nectophore / mature gastrozooid, PalmatGasmat = mature palpon / mature gastrozooid, PneGasmat = Pneumatophore / mature gastrozooid.

```{r neg-pos, results='asis', echo=FALSE, message=FALSE, warning=FALSE, verbose= FALSE, fig.width=14, fig.height=10, fig.cap='(ref:neg-pos)', fig.align='center'}

@@

grid.arrange(tree_reference,Changes_raw,Changes_ratio,legend4, bottom="Tissue",widths = c(2,2,0.25),layout_matrix = layout4)

@@

Out of a total of `r edges_tidy_directional %>% .$gene_tree %>% unique %>% length()` final gene trees considered in these analyses, `r edges_tidy_directional %>% filter(change_direction=="Neutral") %>% .$gene_tree %>% unique %>% length()` gene trees contained branches with neutral changes, `r edges_tidy_directional %>% filter(change_direction=="Positive") %>% .$gene_tree %>% unique %>% length()` gene trees contained branches with positive changes, and `r edges_tidy_directional %>% filter(change_direction=="Negative") %>% .$gene_tree %>% unique %>% length()` gene trees contained branches with negative changes. Expression ratio changes across species-equivalent branches from all gene trees is available in supplementary file 9 (note: in this file 'blast hit' is the most frequent blast hit for the gene tree, gene identity should still be confirmed for each gene, particularly in large gene trees).

The vast majority of gene trees contain species-equivalent branches with neutral changes, including a number of transcription factors and morphogenic signalling pathway genes. For example, in ratios of developing and mature gastrozooids, among the relevant species-equivalent branches in the Wnt gene phylogeny, changes are neutral across Wnt3 and Wnt2 branches, with the exception of a slight positive change across the branch leading to a *Diphyes dispar* Wnt2 paralog, suggesting a higher relative expression of this gene in the developing gastrozooid of *D. dispar* (Fig. \@ref(fig:wnts), \@ref(fig:wntplot)). For all zooid ratios, we find very consistent Wnt3 expression patterns with neutral or very small positive or negative changes in expression ratio across the branches. Indeed, across all cnidarians examined, Wnt3 has consistent localised expression at the oral pole, likely playing a role in axis formation and maintenance [@bagaeva2019wnt; @guder2006wnt; @hensel2014lineage; @hobmayer2000wnt; @momose2008maternally; @nawrocki2013expression; @sanders2015patterns]. However, for Pneumatophore/Gastrozooid ratios, some large changes are seen across branches K and L, leading to a *Agalma elegans* and *Nanomia bijuga* Wnt2 paralog respectively, likewise for Palpon/Gastrozooid a large change is seen across branch J, leading to a *Frillagalma vityazi* Wnt2 paralog. In *Nematostella vectensis* and *Hydractinia echinata*, Wnt2 is expressed in the middle of the polyp [@hensel2014lineage; @kusserow2005unexpected]. Without more detailed spatial expression patterns, it is difficult to know whether these differences in expression between species reflect species-specific differences in axial patterning and morphogenesis, such as an expansion or reduction of the expression domain.

@@

We also looked at the `r edges_tidy_directional %>% filter(change < -5 | change > 5) %>% .$gene_tree %>% unique() %>% length()` gene trees with species-relative branches that have very large changes ( 5 < change < -5), representing putative lineage-specific expression patterns. In Frizzled5/8, Wnt4, and homeobox (putatively Hox-B8) gene-trees, for example, we found very large positive changes in ratios of developing gastroozoid/mature gastrozooids across the same species-equivalent branches, J (leading to *Frillagalma vityazi*) and D (leading to Euphysonectae, the clade comprising *A. elegans*, *N. bijuga*, and *F. vityazi*). In Frizzled5/8 for developing nectophores/gastrozooids we also see a large positive change across branch L (leading to *Nanomia bijuga*) and J (leading to *Frillagalma vityazi*) and negative change across branch K (leading to *Agalma elegans*). In ratios of developing nectophores/gastrozooids, we also found very large positive changes in branch J for Wnt-7b-like, Wnt-4, (putatively) Hox-B8 and NKX1-2. Likewise in pneumatophore/gastrozooid ratios, we also find large positive changes in branch J and L for Frizzled5/8, and a negative change across branch K. Across J, we also see positive changes in pneumatophore/gastrozooid in Hox-B8 and NKX1-2. Finally, for branch K, we see large positive changes in Wnt4 and Wnt7b-like. Many of these genes have been shown to have specific expression domains patterning cnidarian bodies [@hensel2014lineage; @kusserow2005unexpected; @ryan2007pre; @sinigaglia2013bilaterian], and further detailed expression analyses of these genes and others within the zooid are required to determine whether the patterns identified here reflect differences in expression domain between species and zooid.

(ref:wnts) Pruned gene tree of Wnt genes in siphonophore species with expression values (TPM10K) at the tips and nodes. Developing nectophore/mature gastrozooid expression in green (above node), developing gastrozooid/mature gastrozooid expression in orange (below node). Changes across the branch are shown for developing gastrozooid/mature gastrozooid (orange). Red circles are duplication nodes and blue circles are speciation nodes.

```{r wnts, echo=FALSE, message=F, warning=FALSE, verbose= FALSE, fig.height=15, fig.width=15, fig.cap='(ref:wnts)', fig.align = 'center'}

@@

knitr::include_graphics("Figures/Wntplot_new.pdf",auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE))

@@

#Positive GO term enrichment
#edges_tidy_directional %>% filter(change >2) %>% .$gene_tree %>% unique() %>% fishertestSBF(.,"BP")

>>>>

---
bibliography: references.bib
csl: council-of-science-editors-author-date.csl
site: "bookdown::bookdown_site"
header-includes:
    - \usepackage{setspace}\onehalfspacing

output:
  bookdown::word_document2:
    toc: FALSE
    fig_caption: yes
---

@@

```{r zooids-definition, echo=FALSE, message=F, warning=FALSE, verbose= FALSE, fig.cap='(ref:zooids-definition)', out.height='0.5\\textheight', out.width='0.5\\textheight', fig.align = 'center'}

@@

Using STF and SBF, combined with classical differential gene expression (DGE) analysis, we conducted DGE analyses among siphonophore zooids, and one specialized tissue (the pneumatophore, a gas filled float), in seven siphonophore species to address three questions. First, we analyzed differences in gene expression within species to identify patterns that define particular zooid types, and gain insight into the genes that may be playing a role in the structure, maintenance, and functioning of zooids. We also examined whether gene expression patterns could distinguish novel, distinct species-specific zooid types within particular species. Second, we compared gene expression between species, using STF followed by linear models to identify zooid- and species-variable genes. And third, we used phylogenetic comparative methods and SBF to identify putative clade/lineage or zooid specific expression patterns.

@@

```{r methods-comp, echo=FALSE, message=F, warning=FALSE, verbose= FALSE, fig.cap='(ref:methods-comp)', out.height='0.5\\textheight', out.width='0.5\\textheight', fig.align = 'center'}

@@

The third component of variation we considered was among zooids/specialized tissues within species. This was the greatest component of variation (Fig. \@ref(fig:pca-diphyes)--\@ref(fig:pca-physalia)). We identified genes that were significantly differentially expressed in particular zooids, and in the pneumatophore, for each of the sampled species, based on pairwise comparisons (supplementary file 1). In each case, significally differentially expressed refers to genes with higher transcript abundance relative to the other zooid. As it is possible for the same gene to be significantly differentially expressed in pairwise comparisons of several zooids, we also identified a subset of genes that were significantly differentially expressed in only one zooid but not in any other zooid within a given species (Fig. \@ref(fig:unique-zooids), supplementary file 2).

@@

In siphonophores, there are several instances of lineage-specific zooid diversification events. We investigated gene expression patterns between the novel zooid type and the hypothesized most closely related zooid type in three species. In *Bargmannia elongata* there are two morphologically distinct gastrozooids, that we termed "white" and "yellow" gastrozooids (Fig. \@ref(fig:novel-zooids)A and \@ref(fig:novel-zooids)B). The "yellow" gastrozooid is larger and darker and occurs as the 7th-10th gastrozooid on the stem [@dunn2005complex].  In the Portuguese man of war, *Physalia physalis*, the gastrozooid is unique compared to other gastrozooids in other species - it has a mouth, but no tentacle, and the basigaster region is greatly reduced [@Totton1960; @Mackie1960]. Meanwhile the tentacle is associated with another zooid, the tentacular palpon (Fig. \@ref(fig:novel-zooids)C) [@BARDI2007;@munro2019morphology; @Totton1960; @haeckel1888]. In *P. physalis*, both the gastrozooid and the tentacular palpon are considered to be subfunctionalized from an ancestral gastrozooid type [@munro2019morphology]. Finally, in *Agalma elegans*, there are thought to be at least two different palpon types: gastric palpons that arise at the base of the peduncle of the gastrozooid, and a palpon called the B-palpon (Fig. \@ref(fig:novel-zooids)D) [@dunn2006evolution]. The distinction between these two types of palpon is based on the location of these zooids - the gastrozooid is typically the last element of each repeating pattern of zooids along the stem (cormidium), but based on the budding sequence, Dunn *et al* proposed that the enlarged B-palpon is the last element in *A. elegans* [@dunn2006evolution]. Each of these cases represents a different type of novelty: in *Bargmannia elongata* the distinction between zooids was made based on size and color but not on obvious differences in function, in *P. physalis* the gastrozooids and tentacular palpons differ structurally and functionally, and finally in *A. elegans*, gastric palpons and B palpons differ only in colony location, development, and possibly size.

@@

pca_data2<-prcomp(t(pcatest2))
pca_data_perc2<-round(100*pca_data2$sdev^2/sum(pca_data2$sdev^2),1)
test_names2<-colnames(pcatest2) %>% str_split("_")
species2<-lapply(test_names2, `[[`, 1) %>% unlist() %>% str_replace("\\."," ")
treatment2<-lapply(test_names2, `[[`, 2) %>% unlist() 
#replicate<-lapply(test_names, `[[`, 3) %>% unlist()
df_pca_data2 <- data.frame(PC1 = pca_data2$x[,1], PC2 = pca_data2$x[,2], species=species2, treatment=treatment2)
plot_pca_ortholog2<-df_pca_data2 %>% ggplot()+geom_point(aes(PC1,PC2, col=treatment, shape=species), size = 3) + 
  scale_shape_manual(values=c("Agalma elegans"=15,"Nanomia bijuga"=0,"Frillagalma vityazi"=16,"Bargmannia elongata"=9)) +
  xlab(paste0("PC1: ",pca_data_perc2[1],"% variance")) +  ylab(paste0("PC2:",pca_data_perc2[2],"% variance")) +
  scale_color_manual(values=c("GasdevGasmat"="#F78508","NecdevGasmat"="#B9C977","PalmatGasmat"="#F4DF1C","PneGasmat"="#8DCCE2"))

@@

(ref:SOFT-plot) Differential gene expression across tissues (i.e., zooids) and species in siphonophores using Speciation Tree Filtering followed by linear models. Panels A & B show results using gene expression as TPM10K values. Panels C & D show results using gene expression as TPM10K ratios (see Methods for details). Panels B & D show the proportion of variance explained by species-variable genes (SVG, red), tissue-variable genes (TVG, blue), and other (Other, gray). In SVGs, the proportion of variance explained by species is two times greater than that explained by tissues, while in TVGs the proportion of variance explained by tissues is two times greater than that explained by species. All SVGs and TVGs are genes for which both species and tissue explain greater than 75% of the variance. A. Expression values of 3174 ortholog genes cluster largely by species (symbols) in the space defined by the two principal components explaining most of the variation (PC1-PC2). B. Out of 3174 ortholog genes, species-variable genes (SVG) explain most of the variance in gene expression values relative to tissue-variable genes (TVG) (SVG = 2125 genes; TVG = 168 genes). C. Expression ratios of 3174 ortholog genes cluster largely by tissue/zooid (colors) in the space defined by the two principal components explaining most of the variation (PC1-PC2). D. Out of 3174 ortholog genes, species-variable genes (SVG) and tissue-variable genes (TVG) explain similar amount of variance in gene expression ratios (SVG = 494 genes; TVG = 349 genes). Abbreviations: Gasdev = developing gastrozooid, Necdev = developing nectophore, Palmat = mature palpon, Pne = pneumatophore, Gasmat = mature gastrozooid.

```{r SOFT-plot, results='asis', echo=FALSE, message=FALSE, warning=FALSE, verbose= FALSE,out.height='0.5\\textheight', out.width='0.5\\textheight', fig.cap='(ref:SOFT-plot)', fig.align='center'}

@@

Most comparative studies of gene expression focus exclusively on strict 1:1 orthologs, requiring an assessment of orthology across all species. We call this type of analysis Speciation Tree Filtering (STF), as it limits analyses to a subset of gene trees with very specific evolutionary histories. Following this approach, we used Orthofinder 2 (v2.4.0) [@emms2019orthofinder] to identify strict orthologs. For all seven species, we identified `r Orthologs1$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r test3 %>% nrow()` orthologs had expression data across all 7 species. In order to increase the number of genes for analysis, we focused on the four best sampled species (*Agalma elegans*, *Bargmannia elongata*, *Frillagalma vityazi*, and *Nanomia bijuga*). Using data from 4 taxa, we identified `r Orthologs2$Orthogroup %>% unique() %>% length()` strict orthologs, of which `r stvariance %>% nrow()` orthologs had expression values in four zooids/tissues: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore.  Using TPM10K values directly, we found that orthologs clustered largely by species rather than zooid/tissue (Fig. \@ref(fig:SOFT-plot)A). Following @breschi2016gene (see methods), we used linear models to identify the proportion of variance that could be explained by species or zooid/tissue (Fig. \@ref(fig:SOFT-plot)B), and found that of `r stvariance  %>% nrow()` total orthologs, `r stvariance %>% filter(class=="SVG") %>% nrow()` species-variable genes (SVG), and `r stvariance %>% filter(class=="TVG") %>% nrow()` zooid/tissue-variable genes (TVG).

@@

For our Speciation Branch Filtering (SBF) analyses (see Fig. \@ref(fig:methods-comp) and methods), we used transcriptome and genome data from `r species_tree$tip.label %>% length()` cnidarian species to generate a total of `r gene_trees %>% length()` gene trees, of which `r gene_trees_ace_calibrated[!is.na(gene_trees_ace_calibrated)] %>% length()` gene trees passed filtering criteria (see Methods for filtering approach). We used Orthofinder 2 [@emms2019orthofinder] to reconcile the species tree and the gene trees to annotate each gene tree node as either a speciation or duplication event. The number of genes represented in these gene trees is shown in Table \@ref(tab:summarize-libraries). The internal nodes on these gene trees represent `r nodes_ace %>% filter(is.na(species) & Ev=="S") %>% nrow()` speciation events and `r nodes_ace %>% filter(is.na(species) & Ev=="D") %>% nrow()` duplication events. Expression values (TPM10K) were mapped to the tips of the gene trees for each zooid/tissue separately, and ancestral expression values were inferred at internal nodes in a maximum likelihood framework. We focused on a subset of zooids and tissues that are common across the sampled species: gastrozooids (developing and mature), nectophores (developing), palpons (mature), and the pneumatophore. The distribution of expression changes along species-equivalent branches are shown in figure \@ref(fig:changes-branches).

As with the STF and linear model analyses, we investigated the impact of counting efficiency on our comparative analyses. We used the TPM10K values at the tips and nodes to calculate expression ratios and calculated changes in expression ratios across a branch for pairs of tissues, using the mature gastrozooid values as the denominator. The distribution of expression ratio changes along species-equivalent branches (branches in the gene tree that correspond to equivalent branches in the species tree, and are descended from speciation events) are shown in figure \@ref(fig:changes-branches2). We found that the variance of change across a given species-equivalent branch is considerably higher for raw TPM10K values as compared with ratios of expression (Figs. \@ref(fig:variance-structure) & \@ref(fig:variance-structure2)). We also identified the number of gene tree branches that had a negative (change <= -2), positive (change >= 2) or neutral (change > -2 and < 2) change in either TPM10K or TPM10K ratios across the branch (Fig.\@ref(fig:neg-pos)). The mean number of species-equivalent branches considered in these analyses are shown in Table \@ref(tab:summarize-branches) -- this value reflects the mean number of branches across each zooid or zooid ratio (due to incomplete sampling, some zooids may have fewer or more branches; as seen in Fig.\@ref(fig:neg-pos)). For each species-equivalent branch, we found more negative and positive branches than neutral branches when we used TPM10K values, while for expression ratios, we found that the majority of branches show no change across the branch. This suggests that as for the STF analyses, counting efficiency has a large impact on expression change and in turn leads to large gene/branch specific signal. For expression ratios, these results indicate that the vast majority of branches show close to 0 (neutral) change across the branch, suggesting that for closely related genes, expression ratios tend not to differ.

(ref:neg-pos) Phylogenetic analysis of differential gene expression across tissues (i.e., zooids) and species in siphonophores using Speciation Branch Filtering. Top panel show the species tree with branch IDs (A-L). Bottom left panel shows number of branches (ordinate) with negative, neutral, or positive (colors) changes in gene expression values (TPM10K) for different tissues/zooids (abscissa) across each branch in the species tree (species-equivalent branches; see text). Bottom right panel shows number of branches (ordinate) with negative, neutral, or positive (colors) changes in gene expression ratios for different tissues/zooids (abscissa) across each branch in the species tree (species-equivalent branches). Neutral change (green) is defined as <= 2 and >= -2, negative change (red) is defined as <-2, and positive change (blue) is defined as > 2. These units correspond to TPM10K or TPM10K ratios/substitutions per site. GasdevGasmat = developing gastrozooid / mature gastrozooid, NecdevGasmat = developing nectophore / mature gastrozooid, PalmatGasmat = mature palpon / mature gastrozooid, PneGasmat = Pneumatophore / mature gastrozooid.

```{r neg-pos, results='asis', echo=FALSE, message=FALSE, warning=FALSE, verbose= FALSE, out.height='0.8\\textheight', out.width='0.8\\textheight', fig.cap='(ref:neg-pos)', fig.align='center'}

@@

#pdf(file = "figures/changes.pdf", width=14, height=10)
#grid.arrange(tree_reference,Changes_raw,Changes_ratio,legend4, bottom="Tissue",widths = c(2,2,0.25),layout_matrix = layout4)
#dev.off()

knitr::include_graphics("Figures/SBF_ExpressionChanges.pdf",auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE))

@@

Out of a total of `r edges_tidy_directional %>% .$gene_tree %>% unique %>% length()` final gene trees considered in these analyses, `r edges_tidy_directional %>% filter(change_direction=="Neutral") %>% .$gene_tree %>% unique %>% length()` gene trees contained branches with neutral changes in expression ratios, `r edges_tidy_directional %>% filter(change_direction=="Positive") %>% .$gene_tree %>% unique %>% length()` gene trees contained branches with positive changes in expression ratios, and `r edges_tidy_directional %>% filter(change_direction=="Negative") %>% .$gene_tree %>% unique %>% length()` gene trees contained branches with negative changes in expression ratios. Expression ratio changes across species-equivalent branches from all gene trees is available in supplementary file 9 (note: in this file 'blast hit' is the most frequent blast hit for the gene tree, gene identity should still be confirmed for each gene, particularly in large gene trees).

The vast majority of gene trees contained species-equivalent branches with neutral changes in expression ratios, including a number of transcription factors and morphogenic signalling pathway genes. For example, in ratios of developing and mature gastrozooids, among the relevant species-equivalent branches in the Wnt gene phylogeny (Wnt gene identity based on @CONDAMINE2019212), changes were neutral across Wnt3 and Wnt2 branches, with the exception of a slight positive change across the branch leading to a *Diphyes dispar* Wnt2 paralog, suggesting a higher relative expression of this gene in the developing gastrozooid of *D. dispar* (Fig. \@ref(fig:wnts), \@ref(fig:wntplot)). For all zooid ratios, we found very consistent Wnt3 expression patterns with neutral or very small positive or negative changes in expression ratio across the branches. Indeed, across all cnidarians examined, Wnt3 has consistent localised expression at the oral pole, likely playing a role in axis formation and maintenance [@bagaeva2019wnt; @guder2006wnt; @hensel2014lineage; @hobmayer2000wnt; @momose2008maternally; @nawrocki2013expression; @sanders2015patterns]. However, for Pneumatophore/Gastrozooid ratios, some large changes were seen across branches K and L, leading to a *Agalma elegans* and *Nanomia bijuga* Wnt2 paralog respectively, likewise for Palpon/Gastrozooid a large change was seen across branch J, leading to a *Frillagalma vityazi* Wnt2 paralog. In *Nematostella vectensis* and *Hydractinia echinata*, Wnt2 is expressed in the middle of the polyp [@hensel2014lineage; @kusserow2005unexpected]. Without more detailed spatial expression patterns, it is difficult to know whether these differences in expression between species reflect species-specific differences in axial patterning and morphogenesis, such as an expansion or reduction of the expression domain.

@@

We also looked at the `r edges_tidy_directional %>% filter(change < -5 | change > 5) %>% .$gene_tree %>% unique() %>% length()` gene trees with species-relative branches that had very large changes ( 5 < change < -5), representing putative lineage-specific expression patterns (see supplementary file 9). In Frizzled5/8, Wnt4, and homeobox (putatively Hox-B8) gene-trees, for example, we found very large positive changes in ratios of developing gastroozoid/mature gastrozooids across the same species-equivalent branches, J (leading to *Frillagalma vityazi*) and D (leading to Euphysonectae, the clade comprising *A. elegans*, *N. bijuga*, and *F. vityazi*). In Frizzled5/8 for developing nectophores/gastrozooids we also saw a large positive change across branch L (leading to *Nanomia bijuga*) and J (leading to *Frillagalma vityazi*) and negative change across branch K (leading to *Agalma elegans*). In ratios of developing nectophores/gastrozooids, we also found very large positive changes in branch J for Wnt-7b-like, Wnt-4, (putatively) Hox-B8 and NKX1-2. Likewise in pneumatophore/gastrozooid ratios, we also found large positive changes in branch J and L for Frizzled5/8, and a negative change across branch K. Across J, we also saw positive changes in pneumatophore/gastrozooid in Hox-B8 and NKX1-2. Finally, for branch K, we saw large positive changes in Wnt4 and Wnt7b-like. Many of these genes have been shown to have specific expression domains patterning cnidarian bodies [@hensel2014lineage; @kusserow2005unexpected; @ryan2007pre; @sinigaglia2013bilaterian], and further detailed expression analyses of these genes and others within the zooid are required to determine whether the patterns identified here reflect differences in expression domain between species and zooid.

(ref:wnts) Gene tree family of siphonophore Wnt genes available in this analysis, additional WntA and Wnt7 gene trees are presented in Figures \@ref(fig:wntAplot), \@ref(fig:wnt7plot). The tree is presented as two subtrees because upstream Orthofinder analyses separated them into two trees. The tips of the tree correspond to distinct Wnt gene lineages present in seven species of Siphonophores. Numbers at the tips correspond to observed gene expression values (TPM10K), and inferred ancestral values at internal nodes. Developing nectophore/mature gastrozooid expression in green (above node), developing gastrozooid/mature gastrozooid expression in orange (below node). Changes across the branch are shown for developing gastrozooid/mature gastrozooid (orange). Red circles are duplication nodes and blue circles are speciation nodes. Full gene trees with all available cnidarian genes are in Figs \@ref(fig:wntplot) and \@ref(fig:wnt4plot). Wnt gene identity based on @CONDAMINE2019212.

```{r wnts, echo=FALSE, message=F, warning=FALSE, verbose= FALSE, fig.cap='(ref:wnts)',out.height='0.8\\textheight', out.width='0.8\\textheight', fig.align = 'center'}

@@

knitr::include_graphics("Figures/Wnt_plot2_new.pdf",auto_pdf = getOption("knitr.graphics.auto_pdf", FALSE))

@@

#Positive GO term enrichment
edges_tidy_directional %>% filter(change >2) %>% .$gene_tree %>% unique() %>% fishertestSBF(.,"BP") %>% readr::write_csv(., "Supplementary_Data/Supplementaryfile10.csv", na = "NA")

edges_tidy_directional %>% filter(change >2) %>% left_join(GO_to_tips) %>% readr::write_csv(., "Supplementary_Data/Supplementaryfile11.csv", na = "NA")
